On Combinatorial Games and de Bruijn Sequences

### 2016-11-10 infdebruijn.tex
.

% Theorem like env.
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{lemma}[theorem]{Lemma}

@@

\begin{proposition}
For a word $w=w(i)$ be the $i$th word in the above enumeration. Let $l$ be the number of leading zeros in $w$ and let $\hat{w}=\rl{b_1(b^*(w))}{l}$. Then, inserting the cycle $\T{\rl{b_1(w)}{s}}_{s=0}^{n-1}$ after the word $\hat{w}_{1..(n-l)}0^l$ in $\T{u(j)}_{j=0}^{i-1}$ yields the sequence $\T{u(j)}_{j=0}^{i}$.
\end{proposition}

\section{Where can I find $w$ as a sub-word of $w^\infty$? (second try...)}

\begin{definition}
For a word $w$, $max(w)$ is the maximal digit in $w$.
\end{definition}

\begin{definition}
A word $u\in \mathbb{N}$ corresponds to $w\in L_n$ if $u$ is a rotation of $w^{\frac{n}{|w|}}$. Note that each $u\in \mathbb{N}$ corresponds to exactly one word $w\in L_n$.
\end{definition}

\begin{proposition}
\label{newforward}
If $w\in L_n$ and $|w|<n$, then $f^*(w)=f(w)=0^{|w|}u$ for some word $u$.
\end{proposition}
\begin{proof}
Write $\frac{n}{|w|}=c>1$. Thus, $f_1(w)=x$, $f_2(x)=0^{(c-1)|w|}x$ and $f_3(0^{(c-1)|w|}x)=0^{c-1|w|}x=f(w)$. Since $|x|=|w|$, $|f(w)|=n$ and hence, $f^*(w)=f(w)$. In addition, $(c-1)|w|\geq|w|$. 
\end{proof}

\begin{proposition}
\label{1-back}
If $|w|<n$ and $w=w'k$ where $0<k=max(w)$, then $b(w)=uw$ and $max(u)\leq max(w)$. 
\end{proposition}
\begin{proof}
Write $w=w'k$. Thus, $b_1(w)=xk(w'k)^r$, $r>0$. $b_2(xk(w'k)^r)=y(w'k)^r$. $b_3(y(w'k)^r)=uw'k=uw$. It is easy to see that $\max{u}\leq k$.
\end{proof}

\begin{proposition}
\label{*-back}
If $w\in L_n$, $|w|^m=n$, $m>1$ and $w= 0^l\sigma\hat{w}$ such that $\sigma\neq 0$, then $b^*(w)=u\hat{w}w^{m-1}$ for some $u$. Note that in particular, $|b^*(w)|=n$.
\end{proposition}

\begin{proof}
Since $w\in L_n$ and $w\neq 0$, $b_1(w)=w^m=(0^l\sigma w'k)^m$. 
$$b(w)=b_3(b_2((0^l\sigma\hat{w})^m))=b_3((xk)(w'k)^{m-1})=y(w'k)^{m-1}=y\sigma-1\hat{w}w^{m-1}.$$
Now, if $|y\sigma-1\hat{w}w^{m-1}=n$, then $y\sigma-1\hat{w}w^{m-1}=b^*(w)$ and we are done. Otherwise, 
$$b^*(w^{m-1})=b^r(y\sigma-1\hat{w}w^{m-1}).$$
Hence, the words $b(y\sigma-a\hat{w}w^{m-1}),b^2(y\sigma-\hat{w}w^{m-1}),\dots,b^{r-1}(y\sigma-1\hat{w}w^{m-1})$ are of length less than $n$. We apply Proposition \ref{1-back} on these words and we get that for each $l\leq r-1$,  $b^l(y\sigma-1\hat{w}w^{m-1})=u_1u_2\dots u_ly\sigma-1\hat{w}w^{m-1}$.
\end{proof}

\begin{proposition}
Assume that $u\in \mathbb{N}^n$ corresponds to $w\in L_n$ such that $|w|<n$. then, $u$ is a subword of $w^\infty$. 
\end{proposition}

\begin{proof}
If $u=0^n$, then $u$ is a prefix of $w^\infty$ and we are done. Otherwise, $w=0^l\sigma\hat{w}$ where $\sigma\neq 0$ and take $m$ such that $|w|^m=n$. Note that $m>1$. By Propositions \ref{newforward} and  \ref{*-back}, $b^*(w)wf^*(w)=x\hat{w}w^{m-1}w0^|w|y$ is a subword of $w^\infty$. Hence, 
$$\hat{w}(0^l\sigma\hat{w})^m0^l \text{ is a subword of } w^\infty.$$
$u$ is a rotation of $w^m$ thus $u$ is a subword of $\hat{w}(0^l\sigma\hat{w})^m0^l$ which implies that $u$ is a subword of $w^\infty$.
\end{proof}

\begin{proposition}
Assume that $u=yx\in\mathbb{N}^n$ corresponds to $w=xy\in L_n$ where $|w|=n$. If $x\neq 0^r$, then $u$ is a subword of $w^\infty$. 
\end{proposition}

\begin{proof}
We show that $u=yx$ is a subword of $b^*(w)w$. Write $x=0^l\sigma z$ where $\sigma\neq 0$. Thus, since $|w|=n$, $b(w)=\sigma-1zy$. If $b(w)=b^*(w)$, then 
$$b^*(w)w=v(\sigma-1)zyx$$
and we get that $u$ is a subowrd of $b^*(w)b(w)$. Otherwise, $|(\sigma-1)zy|$ does not divides $n$, and in particular, $|\sigma-1zy|<n$. By applying Proposition \ref{1-back} several times, we get that $b*(w)=v(\sigma-1)zy$ for some $v$, and $u=yx$ is a subword of $b^*(w)w=v(\sigma-1)zyxy$.
\end{proof}

\begin{lemma}
\label{z1}
Assume that $w=0^lv\in L_n$ and $|w|=n$. Write $w=0^lz_1\sigma z_2$ where $\sigma$ is the first digit in such that $0^{l+|z_1|}(\sigma+1)z_2$ is lexicographically maximal among its rotations. Take $k\in \N$ such that $|z_1|\leq|(\sigma z_2)^k|$. Then, $z_1$ is a suffix of $(\sigma z_2)^k$.
\end{lemma}

\begin{proof}
Assume for a contradiction that $z_1$ is not a suffix of $(\sigma z_1)^k$, and hence there are $\tau\neq \tau'$ in $\N$ and a word $y$, such that $\tau'y$ is a suffix of $\sigma z_2$, and 
$$z_1=x\tau y(\sigma z_2)^r, \ (\sigma z_2)^k= x'\tau' y(\sigma z_2)^r.$$
Clearly, $\tau<\tau'$ since otherwise, $\tau'<\tau$, and we get that 
$w=0^lz_1\sigma z_2=0^lx\tau y (\sigma z_2)^{r+1}$. However, if we assume that $\tau'<\tau$, $w'=(\sigma z_2)^r0^lx\tau y$ is lexicographically larger than $w$, in contradiction to $w\in L_n$.
\end{proof}

\begin{corrolary}
\label{z1-suffix}
Assume that $w=0^lv\in L_n$ and $|w|=n$. Write $w=0^lz_1\sigma z_2$ where $\sigma$ is the first digit in such that $0^{l+|z_1|}(\sigma+1)z_2$ is lexicographically maximal among its rotations. Then, there are words $x,y$ such that $z_2=xy$, $w=0^ly(\sigma xy)^{r+1}$ and $z_1=y(\sigma xy)^r$.
\end{corrolary}
\begin{proof}
 This is a consequence of the previous Lemma and the fact that $|0^lz_1|=|x(\sigma z_2)|^m$.
\end{proof}

\begin{lemma}
Let $w=0^ly(x0^ly)^{r+1}$ be an $n$-length word such that $y\notin \{0\}^*$. Then, $w$ is not a key-word. 
\end{lemma}

\begin{proof}
Assume for a contradiction that $w$ is a key-word of length $n$, and take a maximal $t\in \N$ such that $x0^ly=x'(0^ly)^{t+1}$. First, we note that $x'\neq \varepsilon$. Indeed, if $x'=\varepsilon$, then $w=(0^ly)(0^ly)^{t(r+1)}$, a periodic word, and $w$ then is not an key-word.

Now we claim that $|x'|<|y|$.
\end{proof}

\begin{proposition}
Assume that $v0^l\in\mathbb{N}^n$ corresponds to $w=0^lv\in L_n$ where $|w|=n$ and $l>0$. Then, $v0^l$ is a subword of $w^\infty$. 
\end{proposition}

\begin{proof}
Write $w=0^lz_1\sigma z_2$ where $\sigma\in \N$ is the first digit in $w$ so that  $0^{l+|z_1|}(\sigma+1)z_2$ is lexicographically maximal among its rotations. Note that such a digit exists since the last digit in $w$ satisfies this requirement. Hence, $v=z_1\sigma z_2$.

By Corollary \ref{z1-suffix}, $z_2=xy$ and $z_1=y(\sigma xy)^r$. Now, since $|0^{l+|z_1|}(\sigma+1)z_2|=n$ and $0^{l+|z_1|}(\sigma+1)z_2$ is lexicographically maximal among its rotations, $0^{l+|z_1|}(\sigma+1)z_2=(w')^{k+1}$ where $w'\in L_n$. Note that $0^{l+|z_1|}$ is a prefix of $w'$. We consider three possibilities

\begin{description}

\item{Case 1.}  $\sigma z_2\in L_n$. We show that in this case, $v0^l$ is a subword of $b^*(b^*(w'))(b^*(w'))w'$, which is a subword of $w^\infty$.

$b_1(w')=w'^{k+1}=0^{l+|z_1|}(\sigma+1)z_2$. Hence, $b(w')=b_3(b_2(0^{l+|z_1|}(\sigma+1)z_1))=\sigma z_2$. Since $\sigma z_2\in L_n$, $b(w')=b^*(w')=\sigma z_2$ and in particular, $|(\sigma z_2)^{m+1}|=n$ for some $m\in \N$. Observe that $|z_1|\leq |\sigma z_2|^m$ and use Lemma \ref{z1} to conclude that $z_1$ is a suffix of $(\sigma z_2)^m$.

By invoking Proposition \ref{1-back} several times, $b^*(\sigma z_2)= u(\sigma z_2)^m$ for some $u$. Hence, $v0^l=z_1\sigma z_2 0^l$ is a subword of 
$$b^*(b^*(w'))b^*(w')b(w')=u(\sigma z_2)^{m+1}0^{l+|z_1|}x'.$$

\item{Case 2.}

Now, $b_1(\sigma z_2) = (\sigma z_2)^{m+1}$ and by Corollary \ref{z1-suffix}, $z_1$ is a suffix of $(\sigma z_2)^m$. Since the last digit in $(\sigma z_2)$ is not zero, by applying Proposition \ref{1-back} several times (possibly none), we get that $b^*{\sigma z_2}=x(\sigma z_2)^m$. As a result we have,
$$b^*(b^*(w'))(b^*(w'))w'= x(\sigma z_2)^m (\sigma z_2) 0^{l+|z_1|}x'.$$
Therefore, $v0^l=z_1\sigma z_2 0^l$ is a subword of  $b^*(b^*(w'))(b^*(w'))w'$ as required.

\end{description}

Before we deal with the other cases, we note that $b_1(\sigma z_2)=b_1(\sigma xy)=x'y(\sigma xy)^{r+1}$ for some $x'$ that satisfies $|x'|=l>0$.

\begin{description}
\item{Case 2.} $\sigma z_2\notin L_n$  and $x'\neq 0^l$. We show that in this case, $v0^l$ is a subword of $b^*(w')w'$, which is a subword of $w^\infty$.

Recall that $b(w')=\sigma z_2$ which is, by assumption, not a key-word. Since $x'\neq 0^l$, several invocations of Proposition \ref{1-back} imply that $b^*(\sigma z_2)=x''y(\sigma z_2)^{r+1}$. Since $v=z_1\sigma z_2=y(\sigma z_2)^{r+1}$, we get that $v0^l$ is a subword of 
$$b^*(w')w'=x''y(\sigma z_2)^{r+1}0^{l+|z_1}u.$$

\item{Case 3.} $\sigma z_2\notin L_n$  and $x'=0^l$. In this case, $b_1(\sigma z_2)$ = $0^ly(\sigma xy)^{r+1}$. Note that $w=0^ly(x''0^ly)^{r+1}$ and use Lemma \ref{} to achieve a contradiction.

\end{description}

\end{proof}

>>>>

% Theorem like env.
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}

@@

\begin{proposition}
	Let $w=w(i)$ for some $i$ and let $l$ be the number of leading zeros in $w$.
	Then, inserting the cycle $\T{R^{-l-n-1}(w),\dots,R^{-l}(w)}$ to $\T{u(j)}_{j=0}^{i-1}$
	 after the word obtained from $R^{-l}(w)$ by decreasing its first letter by one
	 yields the sequence $\T{u(j)}_{j=0}^{i}$.
\end{proposition}
